The role of aneuploidy in the evolution of cancer drug resistance

### 2023-12-04 EvolutionaryRescue1.tex
made comments throughout

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction. A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments. Chromosomal instability (aneuploidy) occurs frequently in cancer cells and introduces large genetics, suggesting that it may be an important mechanism by which populations of cancer cells to adapt to chemotherapy. To determine when this should be the case, we model a population of cancer cells exposed to chemotherapy that can achieve evolutionary rescue with or without aneuploidy. We use branching process approximations to obtain simple expressions for the probability of evolutionary rescue and the effect of aneuploidy. We find that even when aneuploidy by itself only confers partial drug resistance, it can still be an important stepping stone on the way to evolutionary rescue, allowing even small tumors to produce enough genetic diversity to avoid extinction.
\end{abstract}

@@

@@

The role of chromosomal instability (CIN) in the emergence of cancer has been studied extensively in the past decades \citep{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
One hypothesis is that CIN facilitates tumor genesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer. The deletion of tumor suppression genes can happen in two ways: two point mutations deleting both alleles of the TSG (assuming a diploid genotype), or one point mutation and one chromosomal loss event.
Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations \citep{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumor genesis is highly unlikely~\citep{komarova2004optimal}.

\paragraph{Evolutionary rescue.} Populations adapted to a certain environment are vulnerable to environmental changes, which might cause extinction of the population. Examples of such environmental changes include climate change, invasive species or the onset of drug therapies. Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}. 
\emph{Evolutionary rescue} is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted. It is mathematically equivalent to the problem of crossing of fitness valley \citep{weissman2009rate,weissman2010rate}.
There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \citep{uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}.

Models of evolutionary rescue usually assume that the fitness of the wildtype and mutant are homogeneous in time. An exception was given by \citet{marrec2020adapt}, who modeled the fitness of the wildtype and mutant as time dependent. Evolutionary rescue in a quasi-periodic environment has been investigated by \citet{marrec2023evolutionary} with the observation that stochasticity increases the chance of extinction in comparison to the case with deterministic environmental fluctuations. Additionally, \citet{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficient and population size.
Most models focus on the probability that at least one mutation rescues the population. How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute. Additionally, including a range of possible alleles with different fitness which could avert extinction has been coupled with Fischer's geometric model to analyze the probability of evolutionary rescue \citep{wahl2023evolutionary,martin2007distributions,osmond2020genetic}.
Evolutionary rescue that requires two successive mutations has been investigated using diffusion approximation by \citet{martin2013probability}.

@@

We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$. We assume that both chromosomal missegregation and mutation occur during the process of mitosis. Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$. Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively, which we assume is lower than the division rates, $v\cdot\max{(\lambda_w, \lambda_a)}<\min{(\lambda_w, \lambda_a, \lambda_m)}$.
See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes for $u=0 \,\textbf{(A)}, \Delta_a\ll0\, \textbf{(B)},\Delta_a\approx0\, \textbf{(C)},\Delta_a\gg0\, \textbf{(D)}$.

@@

@@

In our model, \emph{evolutionary rescue} occurs when resistant cells appear and fixate ($m_t \gg 1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.
We assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
We therefore define $p_w$ as the probability that a lineage starting from a single wildtype cell avoids extinction by acquiring drug resistance.
Thus, $N^*=1/p_w$ is the threshold tumor size above which evolutionary rescue is very likely, and the rescue probability is given by 
\begin{equation} \label{eq:rescue_prob} 
\presc = 
1-\left(1-p_w\right)^N \approx
1-\e^{-Np_w} = 
1-e^{-N/N^*} .
\end{equation}
where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not necessarily $N p_w$) is small.
Indeed, when $N<1/p_w$, then the probability for evolutionary rescue is $\presc \approx N p_w$  and when $N > 1/p_w$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size. 
\\
In the Appendix, we use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_w$ in different regimes. 
Substituting these  into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
For these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then rescue will likely occur by a direct resistance mutation in a sensitive cell, such that 
\begin{equation} \label{eq:N_m}
N_m^* \approx \frac{\abs{\Delta_w}}{v\lambda_w} \cdot \frac{\lambda_m}{\Delta_m} .
\end{equation}
Here, $\abs{\Delta_w}/\left(v\lambda_w\right)$ is the ratio of the rate at which wild-type cells are decreasing in number and the rate at which they are mutating.

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of drug resistance. 
The threshold tumor size, $N^*$, can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the log of the aneuploid population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a} \cdot \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (partially tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
The first line describes the case in which aneuploids are still effectively killed by the treatment, but not as quickly as the wild type. 
In the second case, the aneuploids are sufficiently resistant that the size of each aneuploid lineage is expected to remain roughly constant. 
In both of these first two cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants. 
In the third case, aneuploidy provides sufficient resistance for the aneuploid population to re-grow the tumor even without additional resistance mutations.
Note that in this case there is no dependence on the parameters characterizing mutants or their production ($v$, $\lambda_m$, and $\Delta_m$).
Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob,rescue_threshold}). Increasing the the missegregation rate $u$ and aneuploid growth rate $\Delta_a$ reduces the threshold tumor size and facilitates the evolution of drug resistance (\Cref{rescue_threshold}).

@@

Therefore, standing genetic variation will drive adaptation to the drug if $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
In contrast, de-novo aneuploids will have a stronger effect on adaptation if the aneuploidy cost $s$ is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$). Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). We derive $s$ by using the formula $s=\tilde{u}\lambda_w/f$ and obtaining the fraction of aneuploid cancer cells $f$ from \citep{lukow2021chromosomal} who mixed together wildtype and aneuploid cell at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time.

For the parameter values of  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=5\cdot10^{-4}$ the ratio of the threshold population sizes is $\tilde{N}^*_a/N^*_a\approx0.125$ which means that standing genetic variation is the main driver of adaptation even when a relatively small fraction cancer cells are aneuploid in the drug-free environment (in this case $f=20\%$). Furthermore, most tumors display significantly higher levels of aneuploidy such that $f\approx 95\%$ even before cancer treatment. In this case, the population threshold ratio becomes $\tilde{N}^*_a/N^*_a\approx0.0025$ and standing genetic variation is a very significant driver of adaptation to the chemotherapeutic drug.
% TODO PLUG IN REAL VALUES ok

@@

Even when evolutionary rescue occurs and leads to recurrence of the tumor, it may take a long time.
When the expected number of resistant lineages that avoid extinction is small the overall expected recurrence time can be estimated by adding two terms: the mean waiting time for evolutionary rescue--the appearance of a resistant lineage that avoid extinction-- (which we denote the mean evolutionary rescue time) and the expected time for proliferation of that lineage back to the original tumor size, $N$ (which we denote the mean proliferation time). However, when the expected number of resistant lineages is large the population of mutant cells behaves deterministically and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time.

\paragraph{Evolutionary rescue time.}
In Appendix C we derive an approximation for $\tau_1$, the expected rescue time without aneuploidy ($u=0$), and $\tau_2$, the expected rescue time with aneuploidy ($u>0$). \Cref{MeanTimeGrowthAneuploidyPlot} shows the agreement between these approximations and simulation results for intermediate and large tumor sizes.

@@

@@

We have modeled a tumor--a population of cancer cells--exposed to drug treatment that causes the population to decline in size towards potential extinction.
In this scenario, the tumor can be "evolutionary rescued", or escape extinction, via two paths. In the direct path, a sensitive cell acquires a mutation that confers resistance that allows it to rapidly grow. In the indirect path, a sensitive cell first becomes aneuploid, which diminishes the effect of the drug, and then an aneuploid cell acquires a mutation that confers resistance (\Cref{figureAneuploidy}).

Using multitype branching processes, we derived the probability of evolutionary rescue of the population of cancer cells under different scenarios for the effect of aneuploidy, ranging from tolerance to partial resistance.
We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}). 
Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the sensitive growth rate $\Delta_w$, the mutation rate $v$, and the aneuploidy rate $u$.

@@

We find that aneuploidy can have a significant effect on evolutionary rescue (\Cref{rescue_prob}). Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N>10^8$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}. This can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse and are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.

Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 1000 days (\Cref{MeanTimeGrowthAneuploidyPlot}A), % TODO do we see tau=100 here...? OK
this can explain the reappearance of cancer even after initial remission. The theoretical predictions for the mean rescue time for tumors smaller then  $10^8$ is greater then 4 years which is consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}.

Paradoxically, we observe from \Cref{MeanTimeGrowthAneuploidyPlot} that the mean time of a rescue mutation to appear is significatly shorter for the case when $u=0$ when compared to the case $u>0$, however this can be explained by the fact this mean time is conditioned on evolutionary rescue and, as a result, aneuploidy increase the \emph{window of opportunity} in which a rescue mutation could appear thus increasing the mean time as well (\Cref{sampleTrajectories}).

We hypothesized that presence of \emph{standing variation} 
--the existence of a subpopulation of aneuploid cancer cells before therapy begins--can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells. Indeed, we observe that even when a small fraction of the initial tumor is aneuploid, evolutionary rescue is more likely to occur through this existing standing variation, rather then through \emph{de novo} aneuploid cells (\Cref{rescue_denovo}).

We observe from \Cref{KaplanMeierfig} that the probability of the mutant cell population reaching size $N$ is approximately zero before time $\tau_2^r$ which is the recurrence time for the deterministic case. This can be explained as follows: in the deterministic case there is a sufficient number of lineages produced such that there exists a lineage where each descendant will only reproduce and not die; the time it takes for this lineage to reach $N$ is the lower bound for the time of all other lineages to reach $N$ and this time cannot be smaller then $\tau_2^r$ by definition. Given that for small values of $N$ we expect that at most a single lineage will rescue the tumor, this lineage cannot reach $N$ before $\tau_2^r$ for the deterministic case \cref{eq:t2det}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction. A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments. Chromosomal instability (aneuploidy) occurs frequently in cancer cells and introduces large genetics, suggesting that it may be an important mechanism by which populations of cancer cells to adapt to chemotherapy. To determine when this should be the case, we model a population of cancer cells exposed to chemotherapy that can achieve evolutionary rescue with or without aneuploidy. We use branching process approximations to obtain simple expressions for the probability of evolutionary rescue and the effect of aneuploidy. We find that even when aneuploidy by itself only confers tolerance to the drug, i.e., is insufficient for rescue, it can still be an important stepping stone on the way to evolutionary rescue, allowing even small tumors to produce enough genetic diversity to avoid extinction.
\end{abstract}

@@

\daniel{RULE OF THUMB: EVERYTHING IN INTRO SHOULD LEAD TOWARD OUR QUESTION/WORK.}

@@

The role of chromosomal instability (CIN) in the emergence of cancer has been studied extensively in the past decades \citep{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}. \daniel{SO WHAT, WHO CARES? SHOULD BE ABOUT WHETHER CIN IS IMPORTANT IN CANCER. (AND ANSWER BETTER BE YES, OR IF ANSWER IS NO OUR HOOK SHOULD BE THAT MAYBE EVEN IF IT IS IMPORTANT IN GENERAL ITS IMPORTANT IN RESISTANCE.)}
One hypothesis is that CIN facilitates tumorigenesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer. The deletion of tumor suppression genes can happen in two ways: two point mutations deleting both alleles of the TSG (assuming a diploid genotype), or one point mutation and one chromosomal loss event.
Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations \citep{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumorigenesis is highly unlikely~\citep{komarova2004optimal}.

\paragraph{Evolutionary rescue.} Populations adapted to a certain environment are vulnerable to environmental changes, which might cause extinction of the population. Examples of such environmental changes include climate change, invasive species, or the onset of drug therapies. Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}. 
\emph{Evolutionary rescue} is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted. \daniel{It is mathematically equivalent to the problem of crossing fitness valleys \citep{weissman2009rate,weissman2010rate}. CUT?}
There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \citep{uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}.
\daniel{WHICH OF THESE ARE RELEVANT FOR CANCER? WHAT'S BEEN DONE IN THE CONTEXT OF CANCER? OR IS THERE EVEN ANY REASON WHY CANCER IS SPECIAL WHEN IT COMES TO EVOLUTIONARY RESCUE?}

Models of evolutionary rescue usually assume that the fitness of the wildtype and mutant are homogeneous in time. An exception was given by \citet{marrec2020adapt}, who modeled the fitness of the wildtype and mutant as time dependent. Evolutionary rescue in a quasi-periodic environment has been investigated by \citet{marrec2023evolutionary} with the observation that stochasticity increases the chance of extinction in comparison to the case with deterministic environmental fluctuations. Additionally, \citet{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficient and population size. \daniel{CUT ALL OF THE ABOVE IN THIS PAR? WE AREN'T REALLY DEALING WITH CHANGE IN TIME EITHER.}
Most models focus on the probability that at least one mutation rescues the population. How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute. Additionally, including a range of possible alleles with different fitness which could avert extinction has been coupled with Fisher's geometric model to analyze the probability of evolutionary rescue \citep{wahl2023evolutionary,martin2007distributions,osmond2020genetic}.
Evolutionary rescue that requires two successive mutations has been investigated using diffusion approximation by \citet{martin2013probability}.

@@

We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make more precise below. We assume that both chromosomal missegregation and mutation occur during the process of mitosis. Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$. Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively, which we assume is lower than the division rates, $v\cdot\max{(\lambda_w, \lambda_a)}<\min{(\lambda_w, \lambda_a, \lambda_m)}$.
See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes for \textbf{(A)} $u=0$ (no aneuploidy), \textbf{(B)} $\Delta_a < 0$, \textbf{(C)} $\Delta_a\approx0$  and \textbf{(D)} $\Delta_a > 0$.

@@

\daniel{SAY MORE ABOUT NATURAL VALUES OF THE PARAMETERS, TABLE}

@@

Aneuploidy may contribute to evolutionary rescue by either preventing population extinction directly (when aneuploidy confers resistance, $\Delta_a>0$) or by delaying extinction until resistant mutants can be establish (when $0>\Delta_a>\Delta_w$).
\daniel{We assume independence between clonal lineages starting from an initial population of $N$ wildtype cells. NEED TO EXPLAIN WHY THIS IS GOOD APPROXIMATION, NOT ASSUMPTION.} (We check the effect of density-dependent growth on our results below).
We therefore define $p_w$ as the probability that a lineage starting from a single wildtype cell avoids extinction by acquiring drug resistance.
The rescue probability of the whole population is then:
\begin{equation} \label{eq:rescue_prob} 
\presc = 
1-\left(1-p_w\right)^N \approx
1-\e^{-Np_w} \equiv 
1-e^{-N/N^*},
\end{equation}
where $N^* \equiv 1/p_w$ is the threshold tumor size above which evolutionary rescue is very likely.
Specifically, when $N \ll N^*$, the probability for evolutionary rescue is $\presc \approx N / N^* \ll 1$  and when $N > N^*$, it is $\presc \approx 1$.

In the Appendix, we use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_w$ in different regimes. 
Substituting these  into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
For these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is both rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then rescue will likely occur by a direct resistance mutation in a sensitive cell, such that 
\begin{equation} \label{eq:N_m}
N_m^* \approx \frac{\abs{\Delta_w}}{v\lambda_w} \cdot \frac{\lambda_m}{\Delta_m} .
\end{equation}
Here, $\abs{\Delta_w}/\left(v\lambda_w\right)$ is the ratio of the rate at which wild-type cells are decreasing in number and the rate at which they are mutating.

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of drug resistance. 
The threshold tumor size, $N^*$, can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the log of the aneuploid population size during the critical time:
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a} \cdot \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (partially tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
The first line describes the case in which aneuploids are still effectively killed by the treatment, but not as quickly as the wild type. 
In the second case, the aneuploids are sufficiently resistant that the size of each aneuploid lineage is expected to remain roughly constant. 
In both of these first two cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants. 
In the third case, aneuploidy provides sufficient resistance for the aneuploid population to re-grow the tumor even without additional resistance mutations.
Note that in this case there is no dependence on the parameters characterizing mutants or their production ($v$, $\lambda_m$, and $\Delta_m$).
Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob,rescue_threshold}). Increasing the the missegregation rate $u$ and aneuploid growth rate $\Delta_a$ reduces the threshold tumor size and facilitates the evolution of drug resistance (\Cref{rescue_threshold}).

@@

Therefore, standing genetic variation will drive adaptation to the drug if $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
In contrast, de-novo aneuploids will have a stronger effect on adaptation if the aneuploidy cost $s$ is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$). Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). 
\daniel{We derive $s$ by using the formula $s=\tilde{u}\lambda_w/f$ and obtaining the fraction of aneuploid cancer cells $f$ from \citep{lukow2021chromosomal} who mixed together wildtype and aneuploid cell at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time.  SEEMS OUT OF PLACE, BELONGS IN MODEL SECTION.}

For the parameter values of  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=5\cdot10^{-4}$ the ratio of the threshold population sizes is $\tilde{N}^*_a/N^*_a\approx0.125$ which means that standing genetic variation is the main driver of adaptation even when a relatively small fraction cancer cells are aneuploid in the drug-free environment (in this case $f=20\%$). Furthermore, most tumors display significantly higher levels of aneuploidy such that $f\approx 95\%$ even before cancer treatment. In this case, the population threshold ratio becomes $\tilde{N}^*_a/N^*_a\approx0.0025$ and standing genetic variation is a very significant driver of adaptation to the chemotherapeutic drug.
% TODO PLUG IN REAL VALUES ok
\daniel{WHAT ABOUT ANEUPLOIDY VS NO-ANEUPLOIDY TAKING STANDING GENETIC VARIATION INTO ACCOUNT? THIS IS MOST IMPORTANT QUANTITY, RIGHT?}

@@

Even when evolutionary rescue occurs and leads to recurrence of the tumor, it may take a long time.
When the expected number of resistant lineages that avoid extinction is small the overall expected recurrence time can be estimated by adding two terms: the mean waiting time for evolutionary rescue--the appearance of a resistant lineage that avoid extinction-- (which we denote the mean evolutionary rescue time) and the expected time for proliferation of that lineage back to the original tumor size, $N$ (which we denote the mean proliferation time). However, when the expected number of resistant lineages is large the population of mutant cells behaves deterministically and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time.
\daniel{WHAT'S THE INTERESTING REGIME HERE? ALSO NEED TO MENTION THAT TIME IS EVEN A QUESTION EARLY IN THE MANUSCRIPT, LIKE IN THE INTRO}

\paragraph{Evolutionary rescue time.}
In Appendix C we derive an approximation for $\tau_1$, the expected rescue time without aneuploidy ($u=0$), and $\tau_2$, the expected rescue time with aneuploidy ($u>0$). \Cref{MeanTimeGrowthAneuploidyPlot} shows the agreement between these approximations and simulation results for intermediate and large tumor sizes.
\daniel{MAKE NOTATION CONSISTENT WITH PREVIOUS SECTION. BE CLEARER ABOUT CONDITIONING}

@@

\daniel{THIS WHOLE TIME SECTION SEEMS LIKE A SKETCH, NOT A DRAFT. WHAT IS THE MESSAGE?}

@@

We have modeled a tumor--a population of cancer cells--exposed to drug treatment that causes the population to decline in size towards potential extinction.
In this scenario, the tumor can be ``evolutionarily rescued'', or escape extinction, via one of two paths. In the direct path, a sensitive cell acquires a mutation that confers resistance that allows it to rapidly grow. In the indirect path, a sensitive cell first becomes aneuploid, which diminishes the effect of the drug, and then a cell from that aneuploid lineage acquires a mutation that confers resistance (\Cref{figureAneuploidy}).

Using multitype branching processes, we derived the probability of evolutionary rescue of the population of cancer cells under different scenarios for the effect of aneuploidy, ranging from tolerance to partial resistance.
We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}). 
\daniel{Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the sensitive growth rate $\Delta_w$, the mutation rate $v$, and the aneuploidy rate $u$.
DIDN'T NEED A MODEL FOR ANY OF THAT. ANYTHING MORE INTERESTING?}

@@

We find that aneuploidy can have a significant effect on evolutionary rescue (\Cref{rescue_prob}). Aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N>10^8$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}. \daniel{This can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse and are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}. THIS IS A GOOD POINT!}

Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 1000 days (\Cref{MeanTimeGrowthAneuploidyPlot}A), % TODO do we see tau=100 here...? OK
this can explain the reappearance of cancer even after initial remission. The theoretical predictions for the mean rescue time for tumors smaller then  $10^8$ is greater then 4 years which is consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}.
\daniel{WHAT ABOUT DISTRIBUTIONS OF RECURRENCE TIME? CAN RAISE THE QUESTION EVEN IF WE DON'T HAVE ANY RESULTS}

\daniel{Paradoxically, we observe from \Cref{MeanTimeGrowthAneuploidyPlot} that the mean time of a rescue mutation to appear is significantly shorter for the case when $u=0$ when compared to the case $u>0$, however this can be explained by the fact this mean time is conditioned on evolutionary rescue and, as a result, aneuploidy increase the \emph{window of opportunity} in which a rescue mutation could appear thus increasing the mean time as well (\Cref{sampleTrajectories}). BETTER TO JUST BE CLEAR AND AVOID PARADOXES IN THE FIRST PLACE}

\daniel{We hypothesized that presence of \emph{standing variation} 
--the existence of a subpopulation of aneuploid cancer cells before therapy begins--can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells. Indeed, we observe that even when a small fraction of the initial tumor is aneuploid, evolutionary rescue is more likely to occur through this existing standing variation, rather then through \emph{de novo} aneuploid cells (\Cref{rescue_denovo}). IF RESCUE PRIMARILY OCCURS THROUGH SGV, THAT SHOULD BE THE CASE WE FOCUS ON, WITH DE NOVO AS THE LITTLE EXTRA SECTION}

\daniel{We observe from \Cref{KaplanMeierfig} that the probability of the mutant cell population reaching size $N$ is approximately zero before time $\tau_2^r$ which is the recurrence time for the deterministic case. This can be explained as follows: in the deterministic case there is a sufficient number of lineages produced such that there exists a lineage where each descendant will only reproduce and not die; the time it takes for this lineage to reach $N$ is the lower bound for the time of all other lineages to reach $N$ and this time cannot be smaller then $\tau_2^r$ by definition. Given that for small values of $N$ we expect that at most a single lineage will rescue the tumor, this lineage cannot reach $N$ before $\tau_2^r$ for the deterministic case \cref{eq:t2det}. DOESN'T BELONG IN DISCUSSION. MAYBE IN RESULTS?}
